Whole-body connectome of a segmented annelid larva

### 2024-03-16 Manuscript.tex
add content to public repo

---
title: "Whole-body connectome of a segmented annelid larva"
author: 
  - name: "Gáspár Jékely<sup>1,2,*</sup>, Csaba Verasztó<sup>2,3</sup>, Sanja Jasek<sup>1,2</sup>, Martin Gühmann<sup>4</sup>, Luis Alberto Bezares Calderón<sup>2</sup>, Elizabeth A. Williams<sup>2</sup>, Réza Shahidi<sup>2,5</sup>"
    affiliation: " <sup>1</sup>Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany <br>  <sup>2</sup>Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom <br>  <sup>3</sup>École Polytechnique Fédérale de Lausanne (EPFL), Lausanne, Switzerland <br> <sup>4</sup>School of Biological Sciences, University of Bristol, Bristol, United Kingdom <br> <sup>5</sup>Electron Microscopy Core Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany <br>  <sup>*</sup>Correspondence: gaspar.jekely @ cos.uni-heidelberg.de"
output:
  pdf_document:
    latex_engine: lualatex
    template: default
  word_document: default
  html_document:
    theme: lumen
geometry: left=0.8in,right=0.8in,top=0.8in,bottom=0.5in
mainfont: Arial
monofont: Times New Roman
documentclass: article
classoption: onecolumn
linestretch: 1.2
fontsize: 10pt
bibliography: references.bib
csl: elife.csl
editor_options: 
  markdown: 
    wrap: sentence
---

```{r include=FALSE}
source("code/statistics_for_paper.R")

```

# Abstract

Nervous systems coordinate effectors across the body during movements.
We know little about the cellular-level structure of synaptic circuits for such body-wide control.
Here we describe the whole-body synaptic connectome of a segmented larva of the marine annelid *Platynereis dumerilii*.
We reconstructed and annotated over 9,000 neuronal and non-neuronal cells in a whole-body serial electron microscopy dataset.
Differentiated cells were classified into 202 neuronal and 92 non-neuronal cell types.
We analyse modularity, multisensory integration, left-right and intersegmental connectivity and motor circuits for ciliated cells, glands, pigment cells and muscles.
We identify several segment-specific cell types, demonstrating the heteromery of the annelid larval trunk.
At the same time, segmentally repeated cell types across the head, the trunk segments and the pygidium suggest the serial homology of all segmental body regions.
We also report descending and ascending pathways, peptidergic circuits and a multi-modal mechanosensory girdle.
Our work provides the basis for understanding whole-body coordination in an entire segmented animal.

# Introduction

Nervous systems coordinate behaviour, physiology and development through synaptic and neuroendocrine signalling.
Signalling occurs specifically between groups of cells, organised into multilayered networks with precise synaptic and neuromodulatory connectivity [@Bentley2016].
Mapping such synaptic and chemical networks in blocks of neural tissue is the central aim of cellular-level connectomics [@Helmstaedter2013; @Deng2019; @Morgan2013; @williams2017].
For synaptic networks, connectomics requires volume imaging by serial electron microscopy (serial EM) [@Schlegel2017].
The comprehensive analysis of whole-body coordination of actions by synaptic circuits would benefit from the cellular-level mapping of entire nervous and effector systems.
Whole-animal synaptic connectomes have so far only been described for the nematode *Caenorhabditis elegans* [@white1986structure; @cook2019whole] and the tadpole larva of the ascidian *Ciona intestinalis* [@ryan2016cns].
Circuits spanning the entire central nervous system (CNS) have also been reconstructed in the larval CNS of the fruit fly *Drosophila melanogaster* [@carreira2018mdn; @miroschnikow2018convergence; @ohyama2015multilevel] and in the three-day-old larva of the annelid *Platynereis dumerilii* [@bezares-calderon2018; @Randel2015; @veraszto2017].
Recently, whole-brain connectomics has become possible in the larval and adult fly brain [@zheng2018complete; @franconville2018building; @winding2023connectome].

Here we report the complete synaptic connectome and the cell-type complement of a three-day-old larva (nectochaete stage) of the marine annelid *Platynereis dumerilii*.
This larval stage has three trunk segments, adult and larval eyes, segmental ciliary bands and a well-developed somatic musculature [@jasek2022].
The larvae show several behaviours, including visual phototaxis [@randel2014neuronal], UV avoidance [@veraszto2018], a startle response [@bezares-calderon2018] and coordinated ciliary activity [@veraszto2017].
Three-day-old larvae do not yet have sensory palps and other sensory appendages (cirri), they do not feed and lack visceral muscles and an enteric nervous system [@brunet2016evolutionary; @williams2015myoinhibitory].

In *Platynereis* larvae, it has been possible to integrate behaviour with synapse-level maps, transgenic labelling of individual neurons, activity imaging and gene knockouts [@bezares-calderon2018; @veraszto2017; @veraszto2018; @jokura2023nitric].
Cellular-resolution gene expression atlases have also been developed for different larval stages.
These can increasingly be integrated with single-cell transcriptomic atlases and synaptic circuit maps [@Achim2015; @vergara2017whole; @williams2017; @Vergara2021].
The registration of a gene expression atlas to a whole-body low-resolution EM volume in the six-day-old *Platynereis* larva has also been reported [@Vergara2021].

We previously reported synaptic connectomes for several whole-body circuits from a transmission EM  volume of a three-day-old larva [@Randel2015].
These include the visual, startle, ciliomotor, nuchal organ, and neurosecretory systems [@bezares-calderon2018; @Randel2015; @shahidi2015serial; @veraszto2017; @veraszto2018; @williams2017].
Here, we report the complete synaptic connectome and the cell-type complement from the same volume of a three-day-old *Platynereis* larva.
Recently we also reported the desmosomal connectome and motoneuron innervation of the somatic musculature [@jasek2022].
The full connectome reconstruction has now allowed us to uncover several new circuits and to consider all circuits in a whole-body context.
We also found several neurons that span the entire length of the larva, highlighting the strength of a whole-body dataset.
These cells and their circuits allow us to generate hypotheses on how whole-body coordination may be achieved by the larval nervous system.
The connectome also allows us to address long-standing hypotheses about the origin of the segmented annelid body-plan and explore patterns of circuit evolution.

# Results

### Serial EM reconstruction of a *Platynereis* larva

We traced and annotated all cells in a previously reported volume EM dataset of a three-day-old (72 hours post fertilisation (hpf)) *Platynereis* larva [@Randel2015] (Figure 1A, Figure 1---figure supplement 1).
The dataset consists of 4,846 layers of 40 nm thin sections scanned by transmission electron microscopy (TEM).
The sections span the entire body of the larva. In the volume, we identified `r num_soma` cells with a soma.

We skeletonised cells containing projections or having an elongated morphology including muscle cells, glia and neurons (Figure 1C, D). During skeletonisation, interconnected nodes are placed in the neurite cross-section profiles of the same neuron across layers. The skeletons are grown until all branches of a neuron have been traced. The individual nodes can be tagged and the skeletons named and multiply annotated. For all neuronal skeletons, we also identified and marked synaptic sites as connectors. Connectors link a node of a presynaptic skeleton to a node of a postsynaptic skeleton partner. Synapses were recognised as presynaptic vesicle clusters at the membrane with no postsynaptic specialisations, as previously described [@randel2014neuronal].
Synapses are monoadic, connecting one presynaptic neuron to one postsynaptic cell. Querying all skeletons and synaptic connectors allowed us to derive the synaptic connectome (Figure 1, Figure 2). We did not identify any gap junctions.

The skeletons across the volume comprised `r with_soma_node_count_sum` nodes and had `r with_soma_presyn_sum` presynaptic and `r with_soma_postsyn_sum` postsynaptic sites.
We could not attach `r num_frag` fragments (`r fragment_node_count_sum` nodes) to a skeleton with a soma.
These fragments contained `r fragment_presyn_sum` presynaptic and `r fragment_postsyn_sum` postsynaptic sites.
Most of the fragments represent short skeletons of twigs (median length `r median_cable_length_frag` µm)(Figure 1---figure supplement 2B) that could not be traced across gaps or low-quality layers.
Overall, `r fraction_fragment`% of all nodes, `r percent_presyn_on_frag`% of presynaptic, and `r percent_postsyn_on_frag`% of postsynaptic sites are on fragments and not assigned to a cell with a soma.

```{r, fig.cap="**Figure 1. Overview of nervous-system anatomy in the three-day-old *Platynereis* larva.** (A) Stylised scanning electron microscopy image of a three-day-old segmented *Platynereis* larva. (B) Morphological rendering of all cells in the EM volume. Spheres represent the position of cell somas. (C) Neurite processes of all neurons in the larva coloured by neuron type (ventral view). Soma are not shown. (D) All effector cells in the larva, including ciliated cells (yellow), glands (grey) and muscles (red). Spheres indicate cell soma (not shown for muscle cells) (E) Summary of cell numbers of different categories in the larva.  (F-H) All presynaptic sites coloured by neuron type shown in (F) ventral, lateral (G), and frontal (H) views. (I) Morphological rendering of all sensory and interneurons on the left side of the larva.  (J) Morphological rendering of all sensory and motor neurons on the left side of the larva. Horizontal lines indicate the position of the cross-sections in (K). (K) Cross-section view of sensory, inter- and motor neuron projections in the ventral nerve cord neuropil. Cross-sections at three antero-posterior positions are shown. Numbers indicate the position of the cross-sections as marked in (J). (L) Position of presynaptic and postsynaptic sites on all sensory neurons of the connectome. (M) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in sensory neurons. (N) Position of presynaptic and postsynaptic sites on all interneurons of the connectome. (O) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in interneurons. (P) Position of presynaptic and postsynaptic sites on all motor neurons of the connectome. (Q) Mean radial synapse density (radius: 1000 nm, centre: soma) of incoming and outgoing synapses in motor neurons. In all anatomical renderings, the outline of the yolk is shown in grey. In (G), the body outline is also shown. Aciculae and chaetae are also shown in grey as segmental markers. Abbreviations: SN, sensory neuron; IN, interneuron; MN, motor neuron, sg0-3, segments 0-3. <br> Figure 1—source data 1. Source data for panels M, O, Q. \n\n"}
#| echo: false 
knitr::include_graphics(path = "Figures/Figure1.png")
```

### Neuroanatomy and neuropils

The *Platynereis* larval nervous system is subdivided into an anterior brain and a rope-ladder-like paired ventral nerve cord (VNC).
The brain and VNC are connected by circumesophageal connectives (Figure 1C, F-H).
The trunk has five segments, an anterior cryptic segment (segment 0)[@steinmetz2011segmental], three main segments (segments 1-3) with parapodia, and a posterior pygidium (Figure 1A, B).

@@

Neurons have a median cable length of `r median_cable_length_all` µm (sensory neurons: `r median_cable_length_SN` µm; interneurons: `r median_cable_length_IN` µm, motoneurons: `r median_cable_length_MN` µm)(Figure 1---figure supplement 2A) and the largest neurons (Ser-tr1 and Loop) a cable length of 1.2-1.7 mm.

Neurons have a median of `r median_presyn_sites` outgoing (presynpatic) and `r median_postsyn_sites` incoming (postsynaptic) synaptic sites with the largest neurons (MC, Ser-tr1, Loop) with over 300 outgoing synapses (Figure 1---figure supplement 2).

@@

The number of outgoing and incoming synaptic sites and skeleton segments scales with the total cable length of skeletons with Pearson's correlation coefficients \>0.7 (Figure 1---figure supplement 2F-H).

Most neuronal trees are unipolar and not highly branched (<100 segments; Figure 1---figure supplement 2H).
Neurons in general do not have primary cilia.
Sensory neurons are bipolar with an axon and a sensory dendrite bearing zero to five sensory cilia at its proximal tip, showing different sensory specialisations [@williams2017; @bezares-calderon2018].
Interneurons are unipolar or pseudo-unipolar.
Ciliomotor neurons have two motor axons emanating from the soma [@veraszto2017], other motoneurons are unipolar.

### Derivation and network analysis of the whole-body synaptic connectome

To define a whole-body synaptic connectome for the larva, we comprehensively identified synapses by traversing the volume and each skeleton multiple times (32,381 synapses). Synapses were connected to the pre- and postsynaptic skeletons (one to one, as synapses are monoadic).
We then retrieved all synapses and their pre- and postsynaptic skeletons and derived a graph (6,725 graph nodes or vertices). From this graph, nodes with less than three connections were removed.
The final synaptic connectome contains `r n_nodes_connectome` nodes (including `r n_fragments` fragments) connected by `r n_edges` directed edges formed by `r n_synapses` in-graph synapses (Figure 2).
The connectome is a sparsely-connected network with a graph density of `r density`.

In the graph, the majority of source nodes (nodes with only outgoing edges) are sensory neurons, sink nodes (nodes with only incoming edges) are mostly effector cells (Figure 2---figure supplement 1). The nodes with the highest degree included ciliomotor neurons (e.g. Loop, Ser-tr1)[@bezares-calderon2018; @veraszto2017], the sensory-motoneuron pygPBunp [@veraszto2017], and the motoneurons of exocrine glands (MNspinning, see below).
Some of the strongest edges (highest number of synapses) were between the pigment-cell motoneuron cioMNcover and prototroch pigment cells, MNspinning and exocrine glands, and the previously described MC cell and ciliated cells [@veraszto2017].

To identify more strongly connected subgraphs within the connectome, we used community detection with the Leiden algorithm [@traag2019louvain].
This analysis combined with force-field-based clustering highlighted several modules (Figures 2).
Nodes in a module are more strongly interconnected than to nodes in other modules (Figure 2---figure supplement 3).
We named the modules based on their primary effector organs or other dominant anatomical characters. We would like to note that this analysis can recover different number of modules depending on the resolution parameter (Figure 2—figure supplement 2C). The selected parameters results in communities that we think best represent functional units.

@@

Querying by two or more annotations allows tallying the number of cells in different category (e.g. number of sensory cells with a single penetrating cilium or the number of gland cells in a particular segment)(Figure 3---figure supplement 3).

@@

### Grouped cell-type-level connectome

@@

### Circuit evolution by duplication and divergence

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  11pt,
]{article}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
  \setmainfont[]{Arial}
  \setmonofont[]{DejaVu Sans Mono}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[lmargin=0.8in,rmargin=0.8in,tmargin=0.8in,bmargin=0.8in]{geometry}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage{lineno}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Whole-body connectome of a segmented annelid larva},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Whole-body connectome of a segmented annelid larva}
\author{}
\date{}

\begin{document}
\maketitle

\linenumbers

\renewcommand*\contentsname{Table of contents}
{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{3}
\tableofcontents
}
\setstretch{1.2}
\hfill\break

Gáspár Jékely\textsuperscript{1,2,*} , Sanja Jasek\textsuperscript{1,2},
Martin Gühmann\textsuperscript{3}, Luis Alberto Bezares
Calderón\textsuperscript{2}, Elizabeth A. Williams\textsuperscript{4},
Réza Shahidi\textsuperscript{5}\\

\textsuperscript{1}Heidelberg University, Centre for Organismal Studies
(COS), 69120 Heidelberg, Germany\\
\textsuperscript{2}Living Systems Institute, University of Exeter,
Exeter, EX4 4QD, United Kingdom\\
\textsuperscript{3}School of Biological Sciences, University of Bristol,
Bristol, United Kingdom\\
\textsuperscript{4}BioSciences, University of Exeter, Stocker Road,
Exeter EX4 4QD, Exeter, UK \textsuperscript{5}Electron Microscopy Core
Facility (EMCF), University of Heidelberg, 69120 Heidelberg, Germany
\textsuperscript{*}Correspondence: gaspar.jekely@cos.uni-heidelberg.de

\section{Abstract}\label{abstract}

Nervous systems coordinate effectors across the body during movements.
We know little about the cellular-level structure of synaptic circuits
for such body-wide control. Here we describe the whole-body synaptic
connectome of a segmented larva of the marine annelid \emph{Platynereis
dumerilii}. We reconstructed and annotated over 9,000 neuronal and
non-neuronal cells in a whole-body serial electron microscopy dataset.
Differentiated cells were classified into 202 neuronal and 92
non-neuronal cell types. We analyse modularity, multisensory
integration, left-right and intersegmental connectivity and motor
circuits for ciliated cells, glands, pigment cells and muscles. We
identify several segment-specific cell types, demonstrating the
heteromery of the annelid larval trunk. At the same time, segmentally
repeated cell types across the head, the trunk segments and the pygidium
suggest the serial homology of all segmental body regions. We also
report descending and ascending pathways, peptidergic circuits and a
multi-modal mechanosensory girdle. Our work provides the basis for
understanding whole-body coordination in an entire segmented animal.

\section{Introduction}\label{introduction}

Nervous systems coordinate behaviour, physiology and development through
synaptic and neuroendocrine signalling. Signalling occurs specifically
between groups of cells, organised into multilayered networks with
precise synaptic and neuromodulatory connectivity (Bentley et al.,
2016). Mapping such synaptic and chemical networks in blocks of neural
tissue is the central aim of cellular-level connectomics (Deng et al.,
2019; Helmstaedter, 2013; Morgan and Lichtman, 2013; Williams et al.,
2017). For synaptic networks, connectomics requires volume imaging by
serial electron microscopy (serial EM) (Schlegel et al., 2017). The
comprehensive analysis of whole-body coordination of actions by synaptic
circuits would benefit from the cellular-level mapping of entire nervous
and effector systems. Whole-animal synaptic connectomes have so far only
been described for the nematode \emph{Caenorhabditis elegans} (Cook et
al., 2019; White et al., 1986) and the tadpole larva of the ascidian
\emph{Ciona intestinalis} (Ryan et al., 2016). Circuits spanning the
entire central nervous system (CNS) have also been reconstructed in the
larval CNS of the fruit fly \emph{Drosophila melanogaster}
(Carreira-Rosario et al., 2018; Miroschnikow et al., 2018; Ohyama et
al., 2015) and in the three-day-old larva of the annelid
\emph{Platynereis dumerilii} (Bezares-Calderón et al., 2018; Randel et
al., 2015; Verasztó et al., 2017). Recently, whole-brain connectomics
has become possible in the larval and adult fly brain (Franconville et
al., 2018; Winding et al., 2023; Zheng et al., 2018).

Here we report the complete synaptic connectome and the cell-type
complement of a three-day-old larva (nectochaete stage) of the marine
annelid \emph{Platynereis dumerilii}. This larval stage has three trunk
segments, adult and larval eyes, segmental ciliary bands and a
well-developed somatic musculature (Jasek et al., 2022). The larvae show
several behaviours, including visual phototaxis (Randel et al., 2014),
UV avoidance (Verasztó et al., 2018), a startle response
(Bezares-Calderón et al., 2018) and coordinated ciliary activity
(Verasztó et al., 2017). Three-day-old larvae do not yet have sensory
palps and other sensory appendages (cirri), they do not feed and lack
visceral muscles and an enteric nervous system (Brunet et al., 2016;
Williams et al., 2015).

In \emph{Platynereis} larvae, it has been possible to integrate
behaviour with synapse-level maps, transgenic labelling of individual
neurons, activity imaging and gene knockouts (Bezares-Calderón et al.,
2018; Jokura et al., 2023; Verasztó et al., 2018, 2017).
Cellular-resolution gene expression atlases have also been developed for
different larval stages. These can increasingly be integrated with
single-cell transcriptomic atlases and synaptic circuit maps (Achim et
al., 2015; Vergara et al., 2021; Vergara et al., 2017; Williams et al.,
2017). The registration of a gene expression atlas to a whole-body
low-resolution EM volume in the six-day-old \emph{Platynereis} larva has
also been reported (Vergara et al., 2021).

We previously reported synaptic connectomes for several whole-body
circuits from a transmission EM volume of a three-day-old larva (Randel
et al., 2015). These include the visual, startle, ciliomotor, nuchal
organ, and neurosecretory systems (Bezares-Calderón et al., 2018; Randel
et al., 2015; Shahidi et al., 2015; Verasztó et al., 2018, 2017;
Williams et al., 2017). Here, we report the complete synaptic connectome
and the cell-type complement from the same volume of a three-day-old
\emph{Platynereis} larva. Recently we also reported the desmosomal
connectome and motoneuron innervation of the somatic musculature (Jasek
et al., 2022). The full connectome reconstruction has now allowed us to
uncover several new circuits and to consider all circuits in a
whole-body context. We also found several neurons that span the entire
length of the larva, highlighting the strength of a whole-body dataset.
These cells and their circuits allow us to generate hypotheses on how
whole-body coordination may be achieved by the larval nervous system.
The connectome also allows us to address long-standing hypotheses about
the origin of the segmented annelid body-plan and explore patterns of
circuit evolution.

\section{Results}\label{results}

\subsubsection{\texorpdfstring{Serial EM reconstruction of a
\emph{Platynereis}
larva}{Serial EM reconstruction of a Platynereis larva}}\label{serial-em-reconstruction-of-a-platynereis-larva}

We traced and annotated all cells in a previously reported volume EM
dataset of a three-day-old (72 hours post fertilisation (hpf))
\emph{Platynereis} larva (Randel et al., 2015) (Figure 1A, Figure
1---figure supplement 1). The dataset consists of 4,846 layers of 40 nm
thin sections scanned by transmission electron microscopy (TEM). The
sections span the entire body of the larva. In the volume, we identified
9,162 cells with a soma.

We skeletonised cells containing projections or having an elongated
morphology including muscle cells, glia and neurons (Figure 1C, D).
During skeletonisation, interconnected nodes are placed in the neurite
cross-section profiles of the same neuron across layers. The skeletons
are grown until all branches of a neuron have been traced. The
individual nodes can be tagged and the skeletons named and multiply
annotated. For all neuronal skeletons, we also identified and marked
synaptic sites as connectors. Connectors link a node of a presynaptic
skeleton to a node of a postsynaptic skeleton partner. Synapses were
recognised as presynaptic vesicle clusters at the membrane with no
postsynaptic specialisations, as previously described (Randel et al.,
2014). Synapses are monoadic, connecting one presynaptic neuron to one
postsynaptic cell. Querying all skeletons and synaptic connectors
allowed us to derive the synaptic connectome (Figure 1, Figure 2). We
did not identify any gap junctions.

The skeletons across the volume comprised 5,661,050 nodes and had 28,717
presynaptic and 27,538 postsynaptic sites. We could not attach 15,020
fragments (896,428 nodes) to a skeleton with a soma. These fragments
contained 4,122 presynaptic and 5,304 postsynaptic sites. Most of the
fragments represent short skeletons of twigs (median length 1.04
µm)(Figure 1---figure supplement 2B) that could not be traced across
gaps or low-quality layers. Overall, 15.8\% of all nodes, 14.4\% of
presynaptic, and 19.3\% of postsynaptic sites are on fragments and not
assigned to a cell with a soma.

\begin{figure}[H]

{\centering \includegraphics[width=1\textwidth,height=\textheight]{Figures/Figure1.png}

}

\caption{\textbf{Figure 1. Overview of nervous-system anatomy in the
three-day-old \emph{Platynereis} larva.} (A) Stylised scanning electron
microscopy image of a three-day-old segmented \emph{Platynereis} larva.
(B) Morphological rendering of all cells in the EM volume. Spheres
represent the position of cell somas. (C) Neurite processes of all
neurons in the larva coloured by neuron type (ventral view). Soma are
not shown. (D) All effector cells in the larva, including ciliated cells
(yellow), glands (grey) and muscles (red). Spheres indicate cell soma
(not shown for muscle cells) (E) Summary of cell numbers of different
categories in the larva. (F-H) All presynaptic sites coloured by neuron
type shown in (F) ventral, lateral (G), and frontal (H) views. (I)
Morphological rendering of all sensory and interneurons on the left side
of the larva. (J) Morphological rendering of all sensory and motor
neurons on the left side of the larva. Horizontal lines indicate the
position of the cross-sections in (K). (K) Cross-section view of
sensory, inter- and motor neuron projections in the ventral nerve cord
neuropil. Cross-sections at three antero-posterior positions are shown.
Numbers indicate the position of the cross-sections as marked in (J).
(L) Position of presynaptic and postsynaptic sites on all sensory
neurons of the connectome. (M) Mean radial synapse density (radius: 1000
nm, centre: soma) of incoming and outgoing synapses in sensory neurons.
(N) Position of presynaptic and postsynaptic sites on all interneurons
of the connectome. (O) Mean radial synapse density (radius: 1000 nm,
centre: soma) of incoming and outgoing synapses in interneurons. (P)
Position of presynaptic and postsynaptic sites on all motor neurons of
the connectome. (Q) Mean radial synapse density (radius: 1000 nm,
centre: soma) of incoming and outgoing synapses in motor neurons. In all
anatomical renderings, the outline of the yolk is shown in grey. In (G),
the body outline is also shown. Aciculae and chaetae are also shown in
grey as segmental markers. Abbreviations: SN, sensory neuron; IN,
interneuron; MN, motor neuron, sg0-3, segments 0-3. Figure 1---source
data 1. Source data for panels M, O, Q.}

\end{figure}%

\subsubsection{Neuroanatomy and
neuropils}\label{neuroanatomy-and-neuropils}

The \emph{Platynereis} larval nervous system is subdivided into an
anterior brain and a rope-ladder-like paired ventral nerve cord (VNC).
The brain and VNC are connected by circumesophageal connectives (Figure
1C, F-H). The trunk has five segments, an anterior cryptic segment
(segment 0)(Steinmetz et al., 2011), three main segments (segments 1-3)
with parapodia, and a posterior pygidium (Figure 1A, B).

@@

Neurons have a median cable length of 97.1 µm (sensory neurons: 101.6
µm; interneurons: 87.77 µm, motoneurons: 139.8 µm)(Figure 1---figure
supplement 2A) and the largest neurons (Ser-tr1 and Loop) a cable length
of 1.2-1.7 mm.

Neurons have a median of 8 outgoing (presynpatic) and 6 incoming
(postsynaptic) synaptic sites with the largest neurons (MC, Ser-tr1,
Loop) with over 300 outgoing synapses (Figure 1---figure supplement 2).

@@

The number of outgoing and incoming synaptic sites and skeleton segments
scales with the total cable length of skeletons with Pearson's
correlation coefficients \textgreater0.7 (Figure 1---figure supplement
2F-H).

Most neuronal trees are unipolar and not highly branched (\textless100
segments; Figure 1---figure supplement 2H). Neurons in general do not
have primary cilia. Sensory neurons are bipolar with an axon and a
sensory dendrite bearing zero to five sensory cilia at its proximal tip,
showing different sensory specialisations (Bezares-Calderón et al.,
2018; Williams et al., 2017). Interneurons are unipolar or
pseudo-unipolar. Ciliomotor neurons have two motor axons emanating from
the soma (Verasztó et al., 2017), other motoneurons are unipolar.

\subsubsection{Derivation and network analysis of the whole-body
synaptic
connectome}\label{derivation-and-network-analysis-of-the-whole-body-synaptic-connectome}

To define a whole-body synaptic connectome for the larva, we
comprehensively identified synapses by traversing the volume and each
skeleton multiple times (32,381 synapses). Synapses were connected to
the pre- and postsynaptic skeletons (one to one, as synapses are
monoadic). We then retrieved all synapses and their pre- and
postsynaptic skeletons and derived a graph (6,725 graph nodes or
vertices). From this graph, nodes with less than three connections were
removed. The final synaptic connectome contains 2,675 nodes (including
467 fragments) connected by 14,066 directed edges formed by 26,881
in-graph synapses (Figure 2). The connectome is a sparsely-connected
network with a graph density of 0.00197.

In the graph, the majority of source nodes (nodes with only outgoing
edges) are sensory neurons, sink nodes (nodes with only incoming edges)
are mostly effector cells (Figure 2---figure supplement 1). The nodes
with the highest degree included ciliomotor neurons (e.g.~Loop,
Ser-tr1)(Bezares-Calderón et al., 2018; Verasztó et al., 2017), the
sensory-motoneuron pygPBunp (Verasztó et al., 2017), and the motoneurons
of exocrine glands (MNspinning, see below). Some of the strongest edges
(highest number of synapses) were between the pigment-cell motoneuron
cioMNcover and prototroch pigment cells, MNspinning and exocrine glands,
and the previously described MC cell and ciliated cells (Verasztó et
al., 2017).

To identify more strongly connected subgraphs within the connectome, we
used community detection with the Leiden algorithm (Traag et al., 2019).
This analysis combined with force-field-based clustering highlighted
several modules (Figures 2). Nodes in a module are more strongly
interconnected than to nodes in other modules (Figure 2---figure
supplement 3). We named the modules based on their primary effector
organs or other dominant anatomical characters. We would like to note
that this analysis can recover different number of modules depending on
the resolution parameter (Figure 2---figure supplement 2C). The selected
parameters results in communities that we think best represent
functional units.

@@

Querying by two or more annotations allows tallying the number of cells
in different category (e.g.~number of sensory cells with a single
penetrating cilium or the number of gland cells in a particular
segment)(Figure 3---figure supplement 3).

@@

\subsubsection{Grouped cell-type-level
connectome}\label{grouped-cell-type-level-connectome}

@@

\subsubsection{Circuit evolution by duplication and
divergence}\label{circuit-evolution-by-duplication-and-divergence}
